### 2025-02-06 latex/paper.tex
project status

\documentclass{article}
\usepackage{hyperref}
\usepackage{float}
\usepackage{url}
% Title Information
\title{Values Encoded in Machine Learning Research}
\author{Isabel Kurth\thanks{Hasso Plattner Institute, Potsdam, Germany.\ \texttt{isabel.kurth@student.hpi.de}}}

% Document Start
\begin{document}

\maketitle
\thispagestyle{plain}

% Abstract Section
\begin{abstract}
% \textbf{Keywords:} Machine Learning, Ethics, Values, Society, Impact
\end{abstract}

% Sections
\section{Introduction}
Machine learning (ML) research has become a pivotal driver of technological advancement, shaping fields as diverse as healthcare, finance, 
and autonomous systems. Despite its transformative potential, the values embedded in influential ML research remain underexamined, raising 
critical questions about whom the field serves and whose needs it prioritizes. Birhane et al. (2023)~\cite{valuesInML2021} underscore that ML research often prioritizes 
performance, generalization, and novelty while neglecting societal needs, ethical considerations, and broader societal impact.

Inspired by this work, our study systematically investigates the values reflected in the 100 most influential papers from NeurIPS 2023 and ICML 2023. 
These conferences represent the cutting edge of ML research and are key arenas where emergent trends and priorities are crystallized. 
By analyzing these highly-cited papers, we aim to uncover patterns in their value commitments, especially regarding how they balance technical rigor 
with societal impact. This work aims to investigate if the most encoded values changed from the years 2008/09 and 2018/19 
(investigated by Birhane et al. (2023)) to more recent papers from 2023.

This paper contributes to the growing discourse on the socio-political dimensions of ML research by providing empirical insights into the values 
upheld in contemporary work. Through our analysis, we aim to encourage researchers, institutions, and policymakers to critically reflect on the 
long-term implications of prioritizing certain values over others.

\section{Methodology}
\subsection{Data collection}
We curated a dataset of the 100 most influential papers from NeurIPS 2023 and ICML 2023 using rankings provided by Paper Digest 
(\url{https://www.paperdigest.org/topic/?topic=nips&year=2023} and \url{https://www.paperdigest.org/topic/?topic=icml&year=2023}). 
These rankings were based on an impact score, which reflects a combination of paper citations, patent citations, etc. It is a score in 1.0-10.0, 
with a higher value indicating a broader impact. We then downloaded the respective papers from the conference website directly 
(\url{https://proceedings.neurips.cc/paper_files/paper/2023} and \url{https://icml.cc/Downloads/2023}). 
The selected papers represent the cutting edge of machine 
learning research and provide a representative sample of current trends and values in the field. We did not use Semantic Scholar as in the Birhane et 
al. (2023) paper because Semantic Scholar did not provide API keys due to high demand. We could also not use Google Scholar to access the citation 
counts because the IP address gets blocked after too many access tries. A list of the PDFs of the 100 papers of both conferences can be found in the 
GitHub repository.

\subsection{Keyword based analysis}
To analyze the values encoded in these papers, we adopted the 74 keywords used by Birhane et al. (2023) in their study of influential ML papers. In their paper
they write that they use 67 values, but in their annotation template for manual annotations (annotations.tsv) they include 74 values. In their results folder they
just mention 62 values. In order to search for as many values as possible we decided to proceed with the 74 values from the annotation file in their data folder.  
These keywords include terms related to performance, generalization, novelty, societal impact, and ethical considerations. 
Unlike Birhane et al., who relied on manual annotation, we developed and deployed an automated keyword-scanning algorithm to systematically identify the presence of 
these keywords across the abstracts, introductions, and conclusions of the selected papers.
The Table\ref{tab:keyword_table} in the Appendix provides a comprehensive list of the values and their associated keywords used in our analysis.
For example we did not only query for the additonal keyword ``Novelty'' but also for ``novel'' as these two
words express the same value and because we use a keyword search, sentences with the word ``novel'' would not be counted towards the keyword ``novelty''.  
For each keyword occurrence, contextual information was extracted to ensure accurate interpretation. 
This semi-automated process balanced the scale of automated analysis with the depth of manual review.

\subsection{Quantitative Summary}
We measured the prevalence of each keyword in the dataset, calculating the relative frequency of each value.

\subsection{Qualitative Summary}
Excerpts containing the keywords were manually reviewed to explore how values are framed and justified. Particular attention was given to the framing of societal impacts and ethical considerations, drawing comparisons with prior findings by Birhane et al.

% Table Example
% \begin{table}[h]
%     \centering
%     \begin{tabular}{lllrrr}
%         \toprule
%         Testing Data & Sources & Hardness & Rule type-1 & Rule type-2 & Rule type-3\\
%         \midrule
%         Sinica & Balanced corpus & Moderate & 92.97 & 94.84 & 96.25 \\
%         Sinorama & Magazine & Difficult & 90.01 & 91.65 & 93.91\\
%         Textbook & Elementary school & Easy & 93.65 & 95.64 & 96.81 \\
%         \bottomrule
%     \end{tabular}
%     \caption{The 50-best oracle performances from the different grammars.}
% \end{table}

% References
\bibliographystyle{plain}
\bibliography{references}

\appendix
\section{Keyword Analysis Table}

\begin{table}[h]
    \centering
    \begin{tabular}{|l|l|}
        \hline
        \textbf{Keywords Birhane Paper} & \textbf{additional keywords for automated keyword search} \\
        \hline
        Novelty & novel \\
        Simplicity & simple \\
        Generalization & generalisable \\
        Flexibility/Extensibility & flexible, flexibility, extensibility, extensible \\
        Robustness & robust \\
        Realistic output & realistic \\
        Formal description/analysis & formal, mathematical \\
        Theoretical guarantees & guarantee \\
        Approximation & approximate \\
        Quantitative evidence (e.g. experiments) & experiment \\
        Qualitative evidence (e.g. examples) & example \\
        Scientific methodology & scientific \\
        Controllability (of model owner) & control \\
        Human-like mechanism & human \\
        Low cost & cheap, cost \\
        Large scale & scale \\
        Generality & general \\
        Principled & principles \\
        Exactness & exact \\
        Preciseness & precise \\
        Concreteness & correct \\
        Automatic & automated \\
        Efficiency & efficient \\
        Building on classic work & classic work \\
        Building in recent work & recent work \\
        Unifying ideas or integrating components & unifying \\
        Identifying limitations & limitations \\
        Critique & criticism \\
        Understanding (for researchers) & understanding \\
        Used in practice/Popular & practice, popular \\
        Reproducibility & reproduce \\
        Easy to implement & implement \\
        Requires few resources & resources \\
        Parallelizability / distributed & parallelizability, parallelization, distributed \\
        Facilitating use (e.g. sharing code) & sharing code \\
        Scales up & scale up \\
        Applies to real world & real world \\
        Learning from humans & humans, learning \\
        Practical & practice \\
        Useful & usefulness \\
        Interpretable (to users) & interpretable \\
        Transparent (to users) & transparent, transparency \\
        Privacy & privacy, private \\
        Fairness & fair \\
        Not socially biased & social bias, socially bias, social, society \\
        User influence & user \\
        Collective influence & collective \\
        Deferral to humans & deferral \\
        Critiqability & criticism \\
        Beneficence & beneficable \\
        Respect for Persons & respect \\
        Autonomy (power to decide) & autonomy, autonome \\
        Explicability & explicable \\
        Respect for Law and public interest & respect for law, respect for public interest \\
        Security & secure \\
        Easy to work with & easy \\
        Realistic world model & world model \\
        Fast & speed \\
        \hline
    \end{tabular}
    \caption{Values and their associated keywords used in the analysis.}
    \label{tab:keyword_table}
\end{table}

\end{document}

### 2025-02-10 latex/paper.tex
updated keywords

@@
 For example we did not only query for the additonal keyword ``Novelty'' but also for ``novel'' as these two
-words express the same value and because we use a keyword search, sentences with the word ``novel'' would not be counted towards the keyword ``novelty''.  
+words express the same value and because we use a keyword search, sentences with the word ``novel'' would not be counted towards the keyword ``Novelty''.  
 For each keyword occurrence, contextual information was extracted to ensure accurate interpretation. 
-This semi-automated process balanced the scale of automated analysis with the depth of manual review.
+This semi-automated process balanced the scale of automated analysis with the depth of manual review. One major limitation to this keyword based approach is 
+that for some keywords this works better than for others and these keywords might then be under- or overrepresented in the analysis. For example for the keyword ``Fairness'' is 
+is really likeli that the keyword search will find the occurences if the paper adresses this topic because they will most liekli use the word in the context. 
+As a counter example the keyword `Èasy to implement'' might not be found as easily because the word ``easy'' is used in many other contexts as well and therefore we can not query for it.
+There are many different words in witch the researchers could express that their model is easy to implement and they are hard to capture all by a simple keyword search.
+In order to keep that limitations as simple as possible
+we try to query for many similar keywords and hope that this will cover most of the values.
+Even tough the keyword search is not perfect, it is a good way to get a first overview of the values encoded in the papers and to get a first impression of the values.
+To stay in the scope of this seminar paper
+we therefor decided to use this approach but we want to explicitly mention this limitation.
 
@@
 We measured the prevalence of each keyword in the dataset, calculating the relative frequency of each value.
+In Figure 1, we present the distribution of the values across all 200 papers from both conferences.
+The top values
+are Performance, Quantitative evidence, Qualitative evidence and Generality.
+We specifically highlited the values that the Brihane et al.
+paper groups as ``User rights'' and ``Ethical Principles'' in the figure.
 
@@
 
+%%%%% muss geupdated werden 
 \begin{table}[h]
@@
         Approximation & approximate \\
-        Quantitative evidence (e.g. experiments) & experiment \\
-        Qualitative evidence (e.g. examples) & example \\
+        Quantitative evidence (e.g. experiments) & quantitative \\
+        Qualitative evidence (e.g. examples) & qualitative \\
         Scientific methodology & scientific \\
@@
         Not socially biased & social bias, socially bias, social, society \\
-        User influence & user \\
+        User influence \\
         Collective influence & collective \\

### 2025-02-13 latex/paper.tex
update

@@
 \usepackage{hyperref}
+\usepackage{booktabs}
 \usepackage{float}
 \usepackage{url}
+\usepackage{graphicx}
+%\usepackage{geometry}
+\usepackage[a4paper, left=0.8in, right=0.8in, top=1in, bottom=1in]{geometry}
+\usepackage{longtable}
+
 % Title Information
@@
 \section{Introduction}
-Machine learning (ML) research has become a pivotal driver of technological advancement, shaping fields as diverse as healthcare, finance, 
-and autonomous systems. Despite its transformative potential, the values embedded in influential ML research remain underexamined, raising 
-critical questions about whom the field serves and whose needs it prioritizes. Birhane et al. (2023)~\cite{valuesInML2021} underscore that ML research often prioritizes 
-performance, generalization, and novelty while neglecting societal needs, ethical considerations, and broader societal impact.
+Machine learning (ML) research has become a driving force behind technological advancements across diverse domains, including healthcare \cite{esteva2019guide}, finance \cite{berg2022fintech}, and autonomous systems \cite{hawkins2021guidance}.
+However, despite its transformative potential, the values embedded in ML research remain underexamined, 
+raising critical questions about whom the field serves and whose needs it prioritizes.
+The previous study by Birhane et al.
+\cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes
+performance, generalization, and novelty while often neglecting ethical considerations, fairness, and broader societal impacts.
 
-Inspired by this work, our study systematically investigates the values reflected in the 100 most influential papers from NeurIPS 2023 and ICML 2023. 
-These conferences represent the cutting edge of ML research and are key arenas where emergent trends and priorities are crystallized. 
-By analyzing these highly-cited papers, we aim to uncover patterns in their value commitments, especially regarding how they balance technical rigor 
-with societal impact. This work aims to investigate if the most encoded values changed from the years 2008/09 and 2018/19 
-(investigated by Birhane et al. (2023)) to more recent papers from 2023.
+In this work, we build upon the foundational analysis by Birhane et al., extending the study to the most influential papers from NeurIPS 2023 and ICML 2023. 
+These conferences represent the cutting edge of ML research and serve as key venues where emerging trends and priorities become crystallized. By systematically analyzing these highly-cited papers, 
+we aim to uncover the prevailing value commitments in ML research and assess whether the field has shifted its focus towards more socially responsible AI.
 
-This paper contributes to the growing discourse on the socio-political dimensions of ML research by providing empirical insights into the values 
-upheld in contemporary work. Through our analysis, we aim to encourage researchers, institutions, and policymakers to critically reflect on the 
-long-term implications of prioritizing certain values over others.
+Our methodology involves a keyword-based analysis, complemented by qualitative reviews, to measure the prevalence of ethical and performance-driven values in recent ML research.
+While previous studies \cite{geyik2019fairness}
+have emphasized the need for fairness-aware ML, our findings indicate that technical considerations such as state-of-the-art (SOTA) performance and computational
+efficiency continue to dominate, often at the expense of user rights and transparency.
+This study contributes to the growing discourse on the socio-political dimensions of ML research by providing empirical insights into the values upheld in contemporary work.
+By situating our findings within
+the broader ML ethics literature, we aim to encourage researchers, institutions, and policymakers to critically reflect on the long-term implications of prioritizing 
+certain values over others.
+Our analysis reveals a pressing need to reorient ML research towards more equitable and inclusive technological development, aligning with calls for ethical AI \cite{floridi2022unified}, \cite{kalluri2020don}
 
@@
 \subsection{Quantitative Summary}
-We measured the prevalence of each keyword in the dataset, calculating the relative frequency of each value.
-In Figure 1, we present the distribution of the values across all 200 papers from both conferences.
-The top values
-are Performance, Quantitative evidence, Qualitative evidence and Generality.
-We specifically highlited the values that the Brihane et al.
-paper groups as ``User rights'' and ``Ethical Principles'' in the figure.
+We measure the prevelence of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference.
+First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
+and lastly we compare the results from the 2023 papers with the results from the 2008/09 and 2018/19 papers from Birhane et al. \cite{valuesInML2021}. 
+In line with the Birhane et al.
+paper we highlight the values that fall under `User rights'.
+These are `Interpretable (To Users)', `Deferral To Humans', `Privacy', `User influence', `Not Socially Biased' and `Fairness'.
+These values are highlighted in all plots in orange.
+The second group of values highlighted are values belonging to `Ethical Principles', These keywords are `Beneficence', `Non-Maleficence', `Respect for Law and Public Interest',
+`Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.
+\subsubsection{ICML and Neurips 2023}
+In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (99\%), accuracy (85\%), 
+Understanding (for researchers) (82.5\%), Efficiency (81\%) and Fast (80\%). Among the values related to user rights and stated in 
+ethical principles the most common ones are `Explicability' (22.5\%) and Privacy (21.5\%).
+
+\begin{figure}[H]
+    \centering
+    %\hspace{-2cm}
+    \includegraphics[width=\textwidth]{../plots/percentage_both_conferences.png}
+    \caption{Prevalence of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023.
+Values related to user rights are highlighted in orange, and values belonging to ethical principles are highlighted in blue.}
+    \label{fig:percentage_both_conferences}
+\end{figure}
+
+By analysing the 100 most influencial paper from both conferences seperately we can see in Figure \ref{fig:percentage_comparison_conferences}
+that the distribution of values are quite similar in both conferences. The top values in both conferences are performance, accuracy, understanding (for researchers), efficiency and fast.
+that the maximum length for a paper is nine pages for Neurips (\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}) and only eight pages for ICML (\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}).
+If the paper
+are shorter the possibility of including values drops due to the space limitation. 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{../plots/percentage_comparison_conferences.png}
+    \caption{Comparison of prevalence of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023.
+Values related to user rights are highlighted in orange, and values belonging to ethical principles are highlighted in blue.}
+    \label{fig:percentage_comparison_conferences}
+\end{figure}
+
+These plots always show the percentage of papers that contain the specific value.
+In order to take into account if one value is detected multiple times in a paper.
+We can see in Figure \ref{fig:total_counts }the total counts of the values in the papers
+for both conferences. This emphasises that Performance is by far the most addressed value in the papers. 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{../plots/histogram_keyword_counts_total_number.png}
+    \caption{Comparison of total counts of the Values.}
+    \label{fig:total_counts}
+\end{figure}
+
+\subsubsection{Comparison 2008/09 and 2018/19 to 2023}
+In Figure \ref{fig:percentage_comparison_years} we compare the distribution of values in the 200 most influential papers from ICML 2023 and NeurIPS 2023 with the results from Birhane et al.
+There are a few values that have a high percentage in both the 2023 papers and the papers from 2008/09 and 2018/19. 
+These are performance (2023: 99\% and Brihane: 96\%), generalisation(77\% and 89\%) and efficiency (81\% and 84\%). 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=\textwidth]{../plots/percentage_comparison_brihane.png}
+    \caption{Comparison of prevelance of values from 2023 to 08/09 and 18/19 from Brihane et al.}
+    \label{fig:percentage_comparison_years}
+\end{figure}
+Strong differences as in values such as `Automatic' and `Progress' can potentially be explaind due to the fact that these 
+words occur often also without specific meaning towards the values.
+
+By looking closer at the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
+we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 expect for `Deferral to humans' and `User influence'. Here the 
+decrease is so small that it could also be due to the keyword search. Especially in the values `Privacy' and `Explicability' we see and strong increase. 
+\begin{figure}[H]
+    \centering
+    \includegraphics[width=0.8\textwidth,height=0.32\textheight]{../plots/subset_percentage_comparison.png}
+    \caption{Comparison of values related to user rights and to ethical principles.}
+    \label{fig:subset_comparison_years}
+\end{figure}
 
 \subsection{Qualitative Summary}
-Excerpts containing the keywords were manually reviewed to explore how values are framed and justified.
-Particular attention was given to the framing of societal impacts and ethical considerations, drawing comparisons with prior findings by Birhane et al.
+In order to investigate if the keyword search was able to capture values correctly, we manually review 
+the sentences flagged for specific values. This qualitative analysis provides a nuanced understanding of how values are encoded in the papers. 
+Table \ref{tab:qualitative_examples_user_influence} provides examples of sentences from the papers that reflect the value `User influence'. These
+examples show that by the automated keywords search in this case sentences get flagged that actually stand in context with the value. 
+\begin{table}[H]
+    \centering
+    \begin{tabular}{p{12cm}}
+        \toprule
+        ``However, in contrast to existing works that target a specific application, without a well defined objective, we propose a more general approach that allows us to unify different user control inputs in a more principled manner.'' \\
+        \midrule
+        ``Controllable Diffusion Models are designed to enable T2I diffusion models to accept more user controls for guiding the generation results.'' \\
+        \midrule
+        ``Recently, a surge of methods have been proposed to gain wider and better user controllability.'' \\
+        \bottomrule
+    \end{tabular}
+    \caption{Random examples of the value \textit{User influence} in the context.}
+    \label{tab:qualitative_examples_user_influence}
+\end{table}
 
-% Table Example
-% \begin{table}[h]
-%     \centering
-%     \begin{tabular}{lllrrr}
-%         \toprule
-%         Testing Data & Sources & Hardness & Rule type-1 & Rule type-2 & Rule type-3\\
-%         \midrule
-%         Sinica & Balanced corpus & Moderate & 92.97 & 94.84 & 96.25 \\
-%         Sinorama & Magazine & Difficult & 90.01 & 91.65 & 93.91\\
-%         Textbook & Elementary school & Easy & 93.65 & 95.64 & 96.81 \\
-%         \bottomrule
-%     \end{tabular}
-%     \caption{The 50-best oracle performances from the different grammars.}
-% \end{table}
+For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly in the first two example sentences.
+In the last case the automated search did wrongly also flag a reference cited in the paper that was publsiehd at the ACM COnference on Fairness, Accountability, and Transparency. 
+In this case the value fairness was not part of the paper but it got flagged. This shows a limitation of the automated appraoch.
+
+\begin{table}[H]
+    \centering
+    \begin{tabular}{p{12cm}}
+        \toprule
+        ``In scenario ( 1), Table 4 shows the fairness issues of GPT-3.5 and GPT-4.'' \\
+        \midrule
+        ``It is unclear whether other properties that have been extensively studied in the literature for the existing public BERT checkpoint, such as robustness, out-of-distribution generalization, fairness or inherent biases, carry over to the crammed model.'' \\
+        \midrule
+        ``In 2022 ACM Confer- ence on Fairness, Accountability, and Transparency , pp.'' \\
+        \bottomrule
+    \end{tabular}
+    \caption{Random examples of the value \textit{Fairness} in the context.}
+    \label{tab:qualitative_examples_fairness}
+\end{table}
+
+We provide in the code the possibility to search for the values the reader is most interested in and one gets 
+as output the names of the paper and the sentences that got flagged.
+
+\section{Discussion}
+Our study extends the analysis of values encoded in machine learning (ML) research by examining the most influential papers from NeurIPS 2023 and ICML 2023.
+In alignment with prior work by Birhane et al.
+(2023),
+we find that performance, accuracy, generalization, and efficiency remain dominant values in contemporary ML research.
+However, our findings also suggest an increasing emphasis on ethical considerations, particularly
+explicability and privacy, indicating a growing awareness of the societal implications of ML advancements.
+A key observation in our study is that the priorities of ML research largely remain entrenched in technical performance metrics.
+The emphasis on state-of-the-art (SOTA) performance and efficiency continues to drive the field,
+yet the explicit inclusion of values related to fairness, transparency, and social responsibility remains relatively limited.
+While there is a slight increase in the prevalence of values associated with user rights, such as
+privacy and explicability, these considerations are still secondary to performance-driven objectives.
+This reinforces concerns raised by Birhane et al.
+(2023) that ML research prioritizes technical achievements over broader
+social impact.
+The keyword-based methodology we employed, while effective for large-scale analysis, introduces limitations regarding contextual accuracy.
+Automated searches for ethical principles may fail to capture nuanced discussions,
+as seen in cases where references to fairness or accountability appeared only in citations rather than substantive discourse.
+This underscores the need for complementary qualitative analysis to provide deeper insights into
+the way these values are operationalized in ML research.
+
+\section{Conclusion}
+Our analysis reveals that while ML research is increasingly acknowledging ethical concerns such as privacy and explicability, technical performance metrics continue to dominate. 
+The persistence of values such as performance, efficiency, and generalization, while important for technological progress, indicates that broader social considerations remain underprioritized. 
+Despite the growing discourse around fairness, transparency, and ethical AI, these values are not yet as deeply embedded in ML research as one might hope.
+
+This study underscores the need for a more deliberate integration of ethical principles into ML research. Future work should explore alternative evaluation metrics that balance technical performance 
+with social responsibility. Moreover, institutions and funding bodies should incentivize research that explicitly addresses societal needs, ensuring that ML advancements contribute to equitable and 
+inclusive technological progress.
+
+Finally, our findings highlight the importance of methodological rigor in value analysis. While keyword-based approaches provide valuable quantitative insights, they must be supplemented with 
+qualitative evaluations to accurately capture the nuances of how values are expressed in ML research.
+As ML continues to shape critical aspects of society, ongoing scrutiny of the values encoded in its
+research will be essential for fostering responsible and equitable AI development.
 
@@
 
+% Appendix
 \appendix
+
 \section{Keyword Analysis Table}
 
-%%%%% muss geupdated werden 
-\begin{table}[h]
-    \centering
-    \begin{tabular}{|l|l|}
+\begin{longtable}{|p{5cm}|p{10cm}|} 
+    \caption{Values and their associated keywords used in the analysis.} \\
     \hline
-        \textbf{Keywords Birhane Paper} & \textbf{additional keywords for automated keyword search} \\
+    \textbf{Keywords Birhane Paper} & \textbf{Additional Keywords for Automated Search} \\
     \hline
-        Novelty & novel \\
-        Simplicity & simple \\
-        Generalization & generalisable \\
-        Flexibility/Extensibility & flexible, flexibility, extensibility, extensible \\
-        Robustness & robust \\
-        Realistic output & realistic \\
-        Formal description/analysis & formal, mathematical \\
-        Theoretical guarantees & guarantee \\
-        Approximation & approximate \\
-        Quantitative evidence (e.g. experiments) & quantitative \\
-        Qualitative evidence (e.g. examples) & qualitative \\
-        Scientific methodology & scientific \\
-        Controllability (of model owner) & control \\
-        Human-like mechanism & human \\
-        Low cost & cheap, cost \\
-        Large scale & scale \\
-        Generality & general \\
-        Principled & principles \\
-        Exactness & exact \\
-        Preciseness & precise \\
-        Concreteness & correct \\
-        Automatic & automated \\
+    \endfirsthead
+
+    \hline
+    \textbf{Keywords Birhane Paper} & \textbf{Additional Keywords for Automated Search} \\
+    \hline
+    \endhead
+
+    % Table content
+    Novelty & novel, innovative, groundbreaking \\ 
+    Simplicity & minimalistic, concise, parsimonious \\ 
+    Generalization & generalisable, generalizable, transferability \\ 
+    Flexibility/Extensibility & flexible, flexibility, extensibility, extensible, adaptable, modular, scalable \\ 
+    Robustness & resilient, fault-tolerant, noise tolerance \\ 
+    Realistic output & authentic, plausible \\ 
+    Formal description/analysis & formal, mathematical, rigorous, analytical, axiomatic, proof-based \\ 
+    Theoretical guarantees & guarantee, provable, convergence proof, theoretical bound, performance bound \\ 
+    Approximation & approximation theory \\ 
+    Quantitative evidence & quantitative, numerical results, empirical study, measurable \\ 
+    Qualitative evidence & case study, illustrative \\ 
+    Scientific methodology & hypothesis-driven, scientific \\ 
+    Controllability (of model owner) & governability, ownership, model steering \\ 
+    Human-like mechanism & biologically inspired, cognitive \\ 
+    Low cost & cheap, cost, affordable, resource-efficient, budget-friendly \\ 
+    Large scale & scalability, big data, high-capacity, massive-scale \\ 
+    Promising & \\ 
+    Generality & broad applicability, domain-independent, versatile \\ 
+    Principled & theoretically sound, axiomatic, methodologically rigorous \\ 
+    Exactness & error-free \\ 
+    Preciseness & high-fidelity \\ 
+    Concreteness & grounded, verifiable \\ 
+    Automatic & self-operating, hands-free \\ 
+    Performance & \\ 
+    Accuracy & precision, recall, F1-score, error rate, reliability \\ 
+    Avoiding train/test discrepancy & train/test, discrepancy, distribution shift, generalization gap \\ 
+    State-of-the-art & SOTA, best performing, cutting-edge, latest \\ 
     Efficiency & efficient \\ 
-        Building on classic work & classic work \\
-        Building in recent work & recent work \\
-        Unifying ideas or integrating components & unifying \\
-        Identifying limitations & limitations \\
-        Critique & criticism \\
-        Understanding (for researchers) & understanding \\
-        Used in practice/Popular & practice, popular \\
-        Reproducibility & reproduce \\
-        Easy to implement & implement \\
-        Requires few resources & resources \\
-        Parallelizability / distributed & parallelizability, parallelization, distributed \\
-        Facilitating use (e.g. sharing code) & sharing code \\
-        Scales up & scale up \\
-        Applies to real world & real world \\
-        Learning from humans & humans, learning \\
-        Practical & practice \\
-        Useful & usefulness \\
-        Interpretable (to users) & interpretable \\
-        Transparent (to users) & transparent, transparency \\
-        Privacy & privacy, private \\
-        Fairness & fair \\
-        Not socially biased & social bias, socially bias, social, society \\
-        User influence \\
-        Collective influence & collective \\
-        Deferral to humans & deferral \\
-        Critiqability & criticism \\
-        Beneficence & beneficable \\
-        Respect for Persons & respect \\
-        Autonomy (power to decide) & autonomy, autonome \\
-        Explicability & explicable \\
-        Respect for Law and public interest & respect for law, respect for public interest \\
-        Security & secure \\
-        Easy to work with & easy \\
-        Realistic world model & world model \\
-        Fast & speed \\
+    Reduced training time & training time, fast training, speed-up, low latency \\ 
+    Memory efficiency & memory-efficient, low memory footprint, RAM optimization \\ 
+    Data efficiency & data-efficient, few-shot, self-supervised, low data regime \\ 
+    Label efficiency & label-efficient, semi-supervised, weak supervision \\ 
+    Energy efficiency & energy-efficient, low power, green AI, sustainable AI \\ 
+    Effectiveness & \\ 
+    Successful & \\ 
+    Building on classic work & classic work, foundational, historical perspective \\ 
+    Building on recent work & recent work, latest advancements, current research \\ 
+    Unifying ideas & unifying, integrative, synergistic, compositional \\ 
+    Identifying limitations & limitations, weaknesses, failure modes \\ 
+    Critique & criticism, critical review \\ 
+    Understanding (for researchers) & understanding, conceptual clarity \\ 
+    Improvement & \\ 
+    Progress & \\ 
+    Used in practice & used in practice, popular, adopted, real-world usage \\ 
+    Reproducibility & reproduce, replication, repeatability, consistent results \\ 
+    Easy to implement & simple to use, straightforward \\ 
+    Requires few resources & resources, low-resource, minimal requirements \\ 
+    Parallelizability & parallelizability, parallelization, distributed \\ 
+    Facilitating use & sharing code, open-source \\ 
+    Scales up & scale up, expands, large-scale deployment \\ 
+    Applies to real world & real world, practical application, real-world relevance \\ 
+    Learning from humans & human learning, human-in-the-loop, interactive learning \\ 
+    Practical & applied AI \\ 
+    Useful & \\ 
+    Interpretable (to users) & interpretable, explainable \\ 
+    Transparent (to users) & transparent, transparency, accountability \\ 
+    Privacy & privacy, private, confidentiality, data protection \\ 
+    Fairness & equitable, bias mitigation \\ 
+    Not socially biased & social bias, socially biased, fairness-aware, bias-free, equitable AI \\ 
+    User influence & user impact, user effect, human influence, user control, user agency \\ 
+    Collective influence & collective, group influence, crowd dynamics, social influence, peer effects \\ 
+    Deferral to humans & human oversight, human intervention, human in the loop, human-AI collaboration \\ 
+    Critiqability & contestability, scrutability, reviewability \\ 
+    Beneficence & beneficable, welfare, positive impact, well-being, prosocial \\ 
+    Non-maleficence & harm avoidance, ethical AI, AI safety, harm reduction \\ 
+    Justice & equity, bias mitigation, equal treatment, social justice \\ 
+    Respect for Persons & human dignity, respect for individuals, respect for rights, human rights \\ 
+    Autonomy & autonomy, self-determination, independence, user agency, free choice \\ 
+    Explicability & explicable, interpretability, transparency, explainability, understandability \\ 
+    Respect for Law & respect for law, respect for public interest, compliance, regulatory adherence, legal AI, governance \\ 
+    Security & secure, cybersecurity, privacy protection, adversarial robustness, data security \\ 
+    Easy to work with & user-friendly, ease of use \\ 
+    Realistic world model & world model, real-world applicability, realistic simulation, grounded AI, embodied intelligence \\ 
+    Fast & speed, low latency, real-time \\ 
     \hline
-    \end{tabular}
-    \caption{Values and their associated keywords used in the analysis.}
-    \label{tab:keyword_table}
-\end{table}
+\end{longtable}
 

### 2025-02-13 latex/paper.tex
repository url
add abstract

@@
 % \textbf{Keywords:} Machine Learning, Ethics, Values, Society, Impact
+Machine learning (ML) research drives technological progress across various domains, yet the values encoded in this research remain underexplored.
+Prior studies suggest that ML research overwhelmingly prioritizes
+technical performance over ethical considerations, fairness, and broader societal impacts.
+Building upon the work of Birhane et al.\cite{valuesInML2021}, we analyze the value commitments in contemporary ML research by examining
+the most influential papers from NeurIPS 2023 and ICML 2023.
+Using a keyword-based and qualitative analysis, we assess the prevalence of ethical and performance-driven values in these papers.
+Our findings reveal that while
+performance, efficiency, and accuracy continue to dominate, there is an increasing—but still limited—acknowledgment of values related to user rights and ethical principles, particularly privacy and explicability.
+However,
+fairness, transparency, and social responsibility remain secondary concerns.
+This study highlights the need for a more deliberate integration of ethical considerations into ML research and calls for alternative evaluation metrics
+that balance technical advancements with social responsibility.
+By providing empirical insights into the values upheld in contemporary ML research, we contribute to the ongoing discourse on the socio-political dimensions of
+AI development and encourage a reorientation towards more equitable and inclusive technological progress.
 \end{abstract}
@@
 raising critical questions about whom the field serves and whose needs it prioritizes.
-The previous study by Birhane et al.
-\cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes
+Therefore it is cirital to analyse the values encoded in ML research.
+Brey states that technology is not neutral but has embedded values \cite{brey2010values}.
+The previous study by Birhane et al. \cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes 
 performance, generalization, and novelty while often neglecting ethical considerations, fairness, and broader societal impacts.
@@
 counts because the IP address gets blocked after too many access tries. A list of the PDFs of the 100 papers of both conferences can be found in the 
-GitHub repository.
+GitHub repository: \url{https://github.com/IsabelKurth/values-in-ml-research}.
 
@@
 
+\section{Results}
 \subsection{Quantitative Summary}
@@
 
+We choose the papers we include in our analysis based on the impact score provided by Paper Digest.
+This score is based on a combination of paper citations, patent citations, etc.
+What needs to be mentioned here is that
+citations always have a time delay and it is hard to determine just shortly after a conference witch papers will have the greatest influence in the future.
+By including 100 papers per conference, we aim to
+capture the most influencial once but it is possible that some papers that will have a great impact in the future are not included in our analysis.
+Also citations only might not be a sufficient criteria for the influence of a paper.
+
 \section{Conclusion}

### 2025-02-13 latex/paper.tex
high resolution plots and typos fixed

@@
 % \textbf{Keywords:} Machine Learning, Ethics, Values, Society, Impact
-Machine learning (ML) research drives technological progress across various domains, yet the values encoded in this research remain underexplored.
+Machine learning (ML) research drives technological progress across various domains, yet the ethical values encoded in this research remain underexplored.
 Prior studies suggest that ML research overwhelmingly prioritizes
@@
 Machine learning (ML) research has become a driving force behind technological advancements across diverse domains, including healthcare \cite{esteva2019guide}, finance \cite{berg2022fintech}, and autonomous systems \cite{hawkins2021guidance}.
-However, despite its transformative potential, the values embedded in ML research remain underexamined, 
-raising critical questions about whom the field serves and whose needs it prioritizes.
-Therefore it is cirital to analyse the values encoded in ML research.
-Brey states that technology is not neutral but has embedded values \cite{brey2010values}.
-The previous study by Birhane et al. \cite{valuesInML2021} has demonstrated that ML research overwhelmingly prioritizes 
-performance, generalization, and novelty while often neglecting ethical considerations, fairness, and broader societal impacts.
+While these advancements have the potential to reshape industries and society at large, the values underpinning ML research remain insufficiently examined.
+This raises crucial questions about whose interests the field serves and what priorities it implicitly endorses.
+As Brey (2010) argues, technology is never value-neutral; instead, it embodies the social, economic, and political values of its creators and institutional contexts \cite{brey2010values}.
+This perspective is particularly relevant for ML,
+where research priorities are often shaped by objectives such as performance optimization, scalability, and generalization—often at the expense of ethical considerations.
+Birhane et al.
+(2021) provide empirical evidence supporting this
+claim, demonstrating that ML research overwhelmingly emphasizes technical advancements while paying comparatively little attention to fairness, accountability, and broader societal implications \cite{valuesInML2021}.
 
-In this work, we build upon the foundational analysis by Birhane et al., extending the study to the most influential papers from NeurIPS 2023 and ICML 2023. 
-These conferences represent the cutting edge of ML research and serve as key venues where emerging trends and priorities become crystallized. By systematically analyzing these highly-cited papers, 
-we aim to uncover the prevailing value commitments in ML research and assess whether the field has shifted its focus towards more socially responsible AI.
+Building on this foundation, our study extends the analysis of value commitments in ML research by examining the most influential papers from NeurIPS 2023 and ICML 2023.
+These premier conferences shape the trajectory of the field by highlighting emerging trends
+and research priorities.
+By systematically analyzing these high-impact papers, we aim to assess whether the ML community has shifted towards a more socially responsible approach or continues to prioritize technical innovation over
+ethical concerns.
 
-Our methodology involves a keyword-based analysis, complemented by qualitative reviews, to measure the prevalence of ethical and performance-driven values in recent ML research.
-While previous studies \cite{geyik2019fairness}
-have emphasized the need for fairness-aware ML, our findings indicate that technical considerations such as state-of-the-art (SOTA) performance and computational
-efficiency continue to dominate, often at the expense of user rights and transparency.
-This study contributes to the growing discourse on the socio-political dimensions of ML research by providing empirical insights into the values upheld in contemporary work.
-By situating our findings within
-the broader ML ethics literature, we aim to encourage researchers, institutions, and policymakers to critically reflect on the long-term implications of prioritizing 
-certain values over others.
-Our analysis reveals a pressing need to reorient ML research towards more equitable and inclusive technological development, aligning with calls for ethical AI \cite{floridi2022unified}, \cite{kalluri2020don}
+Our methodology employs a combination of keyword-based quantitative analysis and qualitative review to assess the prominence of ethical values in recent ML research.
+While prior studies \cite{geyik2019fairness} have emphasized the
+importance of fairness-aware ML, our findings suggest that performance metrics—such as state-of-the-art (SOTA) accuracy and computational efficiency—continue to dominate the discourse, often sidelining considerations of transparency,
+user rights, and broader social impact.
+
+By situating our findings within the broader discourse on ML ethics, this study contributes empirical insights into the implicit value structures that shape contemporary ML research.
+In doing so, we seek to encourage
+researchers, institutions, and policymakers to critically reflect on the long-term implications of these value hierarchies and to foster a more equitable and ethically grounded approach to AI development.
+Our analysis underscores
+the pressing need to reorient ML research toward more inclusive and socially responsible technological progress, aligning with calls for ethical AI \cite{floridi2022unified}, \cite{kalluri2020don}.
 
@@
 \subsection{Data collection}
-We curated a dataset of the 100 most influential papers from NeurIPS 2023 and ICML 2023 using rankings provided by Paper Digest 
-(\url{https://www.paperdigest.org/topic/?topic=nips&year=2023} and \url{https://www.paperdigest.org/topic/?topic=icml&year=2023}). 
+We curated a dataset of the 100 most influential papers from NeurIPS 2023 and ICML 2023 each using rankings provided by Paper Digest 
+\footnote{\url{https://www.paperdigest.org/topic/?topic=nips&year=2023} and \url{https://www.paperdigest.org/topic/?topic=icml&year=2023}}. 
 These rankings were based on an impact score, which reflects a combination of paper citations, patent citations, etc. It is a score in 1.0-10.0, 
-with a higher value indicating a broader impact. We then downloaded the respective papers from the conference website directly 
-(\url{https://proceedings.neurips.cc/paper_files/paper/2023} and \url{https://icml.cc/Downloads/2023}). 
-The selected papers represent the cutting edge of machine 
-learning research and provide a representative sample of current trends and values in the field. We did not use Semantic Scholar as in the Birhane et 
+with a higher value indicating a broader impact. We then downloaded the respective paper from the conference website directly 
+\footnote{\url{https://proceedings.neurips.cc/paper_files/paper/2023} and \url{https://icml.cc/Downloads/2023}}. 
+We choose to select these specific papers with the goal to represent the cutting edge of machine 
+learning research and provide a representative sample of current trends and values in the field.
+It is discussable if citation counts are the best strategy to determine the influencial papers (see Section \ref{sec:discussion}).
+We did not use Semantic Scholar as in the Birhane et 
 al. (2023) paper because Semantic Scholar did not provide API keys due to high demand. We could also not use Google Scholar to access the citation 
@@
 For example we did not only query for the additonal keyword ``Novelty'' but also for ``novel'' as these two
-words express the same value and because we use a keyword search, sentences with the word ``novel'' would not be counted towards the keyword ``Novelty''.  
-For each keyword occurrence, contextual information was extracted to ensure accurate interpretation. 
+words express the same value. 
+For each keyword occurrence, contextual information (the sentences with the keyword occurence) were extracted to ensure accurate interpretation. 
 This semi-automated process balanced the scale of automated analysis with the depth of manual review. One major limitation to this keyword based approach is 
-that for some keywords this works better than for others and these keywords might then be under- or overrepresented in the analysis. For example for the keyword ``Fairness'' is 
-is really likeli that the keyword search will find the occurences if the paper adresses this topic because they will most liekli use the word in the context. 
-As a counter example the keyword `Èasy to implement'' might not be found as easily because the word ``easy'' is used in many other contexts as well and therefore we can not query for it.
-There are many different words in witch the researchers could express that their model is easy to implement and they are hard to capture all by a simple keyword search.
-In order to keep that limitations as simple as possible
-we try to query for many similar keywords and hope that this will cover most of the values.
+that for some keywords this works better than for others and these keywords might then be under- or overrepresented in the analysis. For example for the keyword ``Fairness'' is really likely that the 
+keyword search will find the occurences if the paper adresses this topic because they will most liekly use the word in the context. 
+As a counter example the keyword `Easy to implement'' might not be found as easily because the word ``easy'' is used in many other contexts as well and therefore we decided not to query for it in order to avoid that sentences are flagged with
+the value that do not contain it.
+There are many different words in which the researchers could express that their model is easy to implement and they are hard to capture all by a simple keyword search.
 Even tough the keyword search is not perfect, it is a good way to get a first overview of the values encoded in the papers and to get a first impression of the values.
 To stay in the scope of this seminar paper
-we therefor decided to use this approach but we want to explicitly mention this limitation.
+we therefore decided to use this approach but we want to explicitly mention this limitation.
 
@@
 \subsection{Quantitative Summary}
-We measure the prevelence of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference.
+We measure the prevelence (keyword present in a paper at least once) of each keyword in the 100 papers from the ICML 2023 conference and the 100 papers from the NeurIPS 2023 conference.
 First we investigate the combined distribution of the keywords in both conferences, second we look at both conferences seperately,
@@
 By analysing the 100 most influencial paper from both conferences seperately we can see in Figure \ref{fig:percentage_comparison_conferences}
-that the distribution of values are quite similar in both conferences. The top values in both conferences are performance, accuracy, understanding (for researchers), efficiency and fast.
-that the maximum length for a paper is nine pages for Neurips (\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}) and only eight pages for ICML (\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}).
-If the paper
-are shorter the possibility of including values drops due to the space limitation. 
+that the distribution of values are quite similar in both conferences. The values performance and accuracy are among the top 5 values for both conferences. 
+One can consider that the maximum length for a paper is nine pages for Neurips\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
+This different limitation in the maximum length of the paper does not seem to have a big impact on the distribution of values in the paper.
 \begin{figure}[H]
@@
 In order to take into account if one value is detected multiple times in a paper.
-We can see in Figure \ref{fig:total_counts }the total counts of the values in the papers
-for both conferences. This emphasises that Performance is by far the most addressed value in the papers. 
+We can see in Figure \ref{fig:total_counts} the total counts of the values in the papers
+for both conferences. This emphasises that `Performance' is by far the most addressed value in the papers. 
 \begin{figure}[H]
@@
 
-By looking closer at the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
-we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 expect for `Deferral to humans' and `User influence'. Here the 
-decrease is so small that it could also be due to the keyword search. Especially in the values `Privacy' and `Explicability' we see and strong increase. 
+Zooming onto the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
+we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 except for `Deferral to humans' and `User influence'. For the latter the 
+decrease is so small that it could also be due to the keyword search. Especially in the values `Privacy' and `Explicability' we see a strong increase. 
 \begin{figure}[H]
@@
 For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly in the first two example sentences.
-In the last case the automated search did wrongly also flag a reference cited in the paper that was publsiehd at the ACM COnference on Fairness, Accountability, and Transparency. 
-In this case the value fairness was not part of the paper but it got flagged. This shows a limitation of the automated appraoch.
+In the last case the automated search also flaged a reference cited in the paper that was published at the ACM Conference on Fairness, Accountability, and Transparency. 
+In this case the value `Fairness' was not part of the paper but it got flagged. This shows a limitation of the automated approach. One could argue that citing a paper that deals with fairness
+also coud be counted as a value in the paper.
 
@@
 \section{Discussion}
+\label{sec:discussion}
 Our study extends the analysis of values encoded in machine learning (ML) research by examining the most influential papers from NeurIPS 2023 and ICML 2023.
@@
 This score is based on a combination of paper citations, patent citations, etc.
-What needs to be mentioned here is that
+It should be mentionedis that
 citations always have a time delay and it is hard to determine just shortly after a conference witch papers will have the greatest influence in the future.
@@
 capture the most influencial once but it is possible that some papers that will have a great impact in the future are not included in our analysis.
-Also citations only might not be a sufficient criteria for the influence of a paper.
+Also citations by themselfs might not be a sufficient criterion for the future influence of a paper.
 
@@
     \hline
+    \label{tab:keyword_table}
 \end{longtable}

### 2025-02-14 latex/paper.tex
restructuring of section analysis

@@
 Unlike Birhane et al., who relied on manual annotation, we developed and deployed an automated keyword-scanning algorithm to systematically identify the presence of 
-these keywords across the abstracts, introductions, and conclusions of the selected papers.
+these keywords.
+Brihane et al.
+did only analyse the sections abstract, introduction, discussion and conclusion of the selected papers.
+We assume that they focused on these parts of the papers because they are most likely to contain the values and to minimize the workload for the manual annotations.
+Because we used an automated approach we decided to scan the whole paper (excluding the References section).
+We exclude this section because otherwise here references get flagged with the value `Fairness' for example with they were pubslished at the ACM Conference on Fairness, Accountability, and Transparency.
+This would result in a false positive.
+
 The Table\ref{tab:keyword_table} in the Appendix provides a comprehensive list of the values and their associated keywords used in our analysis.
@@
 `Respect for Persons', `Autonomy (Power to Decide)', `Explicability' and `Justice'.
+
 \subsubsection{ICML and Neurips 2023}
-In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (99\%), accuracy (85\%), 
-Understanding (for researchers) (82.5\%), Efficiency (81\%) and Fast (80\%). Among the values related to user rights and stated in 
-ethical principles the most common ones are `Explicability' (22.5\%) and Privacy (21.5\%).
+In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (97.5\%), accuracy (75\%), 
+State-of-the-art (71\%), Efficiency (70\%) and Improvement (67.5\%). Among the values related to user rights and stated in 
+ethical principles the most common ones are Privacy (16.5\%) and `Explicability' (15.5\%).
 
@@
 By analysing the 100 most influencial paper from both conferences seperately we can see in Figure \ref{fig:percentage_comparison_conferences}
-that the distribution of values are quite similar in both conferences. The values performance and accuracy are among the top 5 values for both conferences. 
+that the distribution of values are quite similar in both conferences. The values `Performance', `Accuracy' and `State-of-the-art' are among the top 5 values for both conferences. 
 One can consider that the maximum length for a paper is nine pages for Neurips\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
@@
 There are a few values that have a high percentage in both the 2023 papers and the papers from 2008/09 and 2018/19. 
-These are performance (2023: 99\% and Brihane: 96\%), generalisation(77\% and 89\%) and efficiency (81\% and 84\%). 
+These are `Performance' (2023: 97.5\% and Brihane: 96\%), `Generalisation' (65.5\% and 89\%) and `Efficiency' (70\% and 84\%). 
 \begin{figure}[H]
@@
 Zooming onto the values related to user rights and ethical principles in Figure \ref{fig:subset_comparison_years}
-we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 except for `Deferral to humans' and `User influence'. For the latter the 
-decrease is so small that it could also be due to the keyword search. Especially in the values `Privacy' and `Explicability' we see a strong increase. 
+we can see an increase in the percentage of papers with these values from 2008/09 and 2018/19 to 2023 except for `Interpretable (to users)' (minor decrease), `Beneficence' (strong decrease), `Deferral to humans' (low decrease) and `User influence' (minor decrease).
+For the values expect `Beneficence' the decrease is so small that it could be due to the keyword search.
+The decrease for `Beneficence' is really strong from 13\% to only 1.5\%.
+One possible explanation is that it is hard to query by a keyword search for this value.
+Especially in the values `Privacy' (from 3\% to 16.5\%), `Explicability' (from 2\% to 15.5\%) and `Fairness' (from 1\% to 10\%) we see a strong increase.
 \begin{figure}[H]
@@
         \midrule
-        ``Controllable Diffusion Models are designed to enable T2I diffusion models to accept more user controls for guiding the generation results.'' \\
+        ``However, user controllability of the generated image, and fast adaptation to new tasks still remains an open challenge, currently mostly addressed by costly and long retraining and fine-tuning or ad-hoc adaptations to specific image generation tasks.'' \\
         \midrule
@@
 
-For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly in the first two example sentences.
-In the last case the automated search also flaged a reference cited in the paper that was published at the ACM Conference on Fairness, Accountability, and Transparency. 
-In this case the value `Fairness' was not part of the paper but it got flagged. This shows a limitation of the automated approach. One could argue that citing a paper that deals with fairness
-also coud be counted as a value in the paper.
+For the values `Fairness' we can see in Table \ref{tab:qualitative_examples_fairness} that the keyword search was able to capture the values correctly 
+in the first two example sentences.In the last case the automated search also flaged a reference cited in the paper that was published at the 
+ACM Conference on Fairness, Accountability, and Transparency. Here the References section was wrongly inclded in the search for this paper, because
+the automatic exclusion of this section does not work perfectly. In this case the value `Fairness' was not part of the paper but it got flagged. 
+This shows a limitation of the automated approach. One could argue that citing a paper that deals with fairnessalso coud be counted as a value in the paper.
 
@@
         \midrule
-        ``In 2022 ACM Confer- ence on Fairness, Accountability, and Transparency , pp.'' \\
+        ``In 2022 ACM Conference on Fairness, Accountability, and Transparency , pp.'' \\
         \bottomrule
@@
 (2023),
-we find that performance, accuracy, generalization, and efficiency remain dominant values in contemporary ML research.
+we find that performance, accuracy, state-of-the-art, and efficiency remain dominant values in contemporary ML research.
 However, our findings also suggest an increasing emphasis on ethical considerations, particularly
-explicability and privacy, indicating a growing awareness of the societal implications of ML advancements.
+explicability, privacy, and fairness indicating a growing awareness of the societal implications of ML advancements.
 A key observation in our study is that the priorities of ML research largely remain entrenched in technical performance metrics.
@@
 Despite the growing discourse around fairness, transparency, and ethical AI, these values are not yet as deeply embedded in ML research as one might hope. 
+This study underscores the need for a more deliberate integration of ethical principles into ML research.
 
-This study underscores the need for a more deliberate integration of ethical principles into ML research. Future work should explore alternative evaluation metrics that balance technical performance 
-with social responsibility. Moreover, institutions and funding bodies should incentivize research that explicitly addresses societal needs, ensuring that ML advancements contribute to equitable and 
-inclusive technological progress.
+This paper analyses papers from the two big ML conferences ICML and Neurips.
+There are specialized ML conferences dedicated to the study of fairness, accountability and transparency in AI.
+It can be debated whether the fact that these specialized conferences exist ensures that a focus on performance values at ICML and Neurips seems justifiable.
 
@@
     Large scale & scalability, big data, high-capacity, massive-scale \\ 
-    Promising & \\ 
+    Promising & - \\ 
     Generality & broad applicability, domain-independent, versatile \\ 
@@
     Automatic & self-operating, hands-free \\ 
-    Performance & \\ 
+    Performance & - \\ 
     Accuracy & precision, recall, F1-score, error rate, reliability \\ 
@@
     Energy efficiency & energy-efficient, low power, green AI, sustainable AI \\ 
-    Effectiveness & \\ 
-    Successful & \\ 
+    Effectiveness & - \\ 
+    Successful & - \\ 
     Building on classic work & classic work, foundational, historical perspective \\ 
@@
     Understanding (for researchers) & understanding, conceptual clarity \\ 
-    Improvement & \\ 
-    Progress & \\ 
+    Improvement & - \\ 
+    Progress & - \\ 
     Used in practice & used in practice, popular, adopted, real-world usage \\ 
@@
     Practical & applied AI \\ 
-    Useful & \\ 
+    Useful & - \\ 
     Interpretable (to users) & interpretable, explainable \\ 
@@
     Critiqability & contestability, scrutability, reviewability \\ 
-    Beneficence & beneficable, welfare, positive impact, well-being, prosocial \\ 
+    Beneficence & beneficable, welfare, positive impact, well-being, prosocial, altruistic, altruism, social good, ethical principle \\ 
     Non-maleficence & harm avoidance, ethical AI, AI safety, harm reduction \\ 

### 2025-02-17 latex/paper.tex
final updates

@@
 
-\subsubsection{ICML and Neurips 2023}
-In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and Neurips 2023. The top values are performance (97.5\%), accuracy (75\%), 
+\subsubsection{ICML and NeurIPS 2023}
+In Figure \ref{fig:percentage_both_conferences} we show the prevelance of values in all 200 papers combined from ICML and NeurIPS 2023. The top values are performance (97.5\%), accuracy (75\%), 
 State-of-the-art (71\%), Efficiency (70\%) and Improvement (67.5\%). Among the values related to user rights and stated in 
@@
 that the distribution of values are quite similar in both conferences. The values `Performance', `Accuracy' and `State-of-the-art' are among the top 5 values for both conferences. 
-One can consider that the maximum length for a paper is nine pages for Neurips\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
+One can consider that the maximum length for a paper is nine pages for NeurIPS\footnote{\url{https://neurips.cc/Conferences/2023/PaperInformation/NeurIPS-FAQ}} and only eight pages for ICML\footnote{\url{https://icml.cc/Conferences/2023/StyleAuthorInstructions}}.
 This different limitation in the maximum length of the paper does not seem to have a big impact on the distribution of values in the paper.
@@
 
-This paper analyses papers from the two big ML conferences ICML and Neurips.
+This paper analyses papers from the two big ML conferences ICML and NeurIPS.
 There are specialized ML conferences dedicated to the study of fairness, accountability and transparency in AI.
-It can be debated whether the fact that these specialized conferences exist ensures that a focus on performance values at ICML and Neurips seems justifiable.
+It can be debated whether the fact that these specialized conferences exist ensures that a focus on performance values at ICML and NeurIPS seems justifiable.
 
